Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-12-22 manuscripts/laplacian_eigenmaps/arxiv/laplacian_eigenmaps_regression.tex
Added cover letter for Inf and Inf.

% Abstract
\begin{abstract}
	In this paper we study the statistical properties of Principal Components Regression with Laplacian Eigenmaps (PCR-LE), a method for nonparametric regression based on Laplacian Eigenmaps (LE). PCR-LE works by projecting a vector of observed responses ${\bf Y} = (Y_1,\ldots,Y_n)$ onto a subspace spanned by certain eigenvectors of a neighborhood graph Laplacian. We show that PCR-LE achieves minimax rates of convergence for random design regression over Sobolev spaces. Under sufficient smoothness conditions on the design density $p$, PCR-LE achieves the optimal rates for both estimation (where the optimal rate in squared $L^2$ norm is known to be $n^{-2s/(2s + d)}$) and goodness-of-fit testing ($n^{-4s/(4s + d)}$). We also show that PCR-LE is \emph{manifold adaptive}: that is, we consider the situation where $\mc{X}$ is a manifold of small intrinsic dimension $m$, and give upper bounds establishing that PCR-LE achieves the faster minimax estimation ($n^{-2s/(2s + m)}$) and testing ($n^{-4s/(4s + m)}$) rates of convergence. Interestingly, these rates are almost always much faster than the known rates of convergence of graph Laplacian eigenvectors to their population-level limits; in other words, for this problem regression with estimated features appears to be much easier, statistically speaking, than estimating the features itself. We support these theoretical results with empirical evidence.
\end{abstract}

>>>>

% Abstract
\begin{abstract}
	In this paper we study the statistical properties of Principal Components Regression with Laplacian Eigenmaps (PCR-LE), a method for nonparametric regression based on Laplacian Eigenmaps (LE). PCR-LE works by projecting a vector of observed responses ${\bf Y} = (Y_1,\ldots,Y_n)$ onto a subspace spanned by certain eigenvectors of a neighborhood graph Laplacian. We show that PCR-LE achieves minimax rates of convergence for random design regression over Sobolev spaces. Under sufficient smoothness conditions on the design density $p$, PCR-LE achieves the optimal rates for both estimation (where the optimal rate in squared $L^2$ norm is known to be $n^{-2s/(2s + d)}$) and goodness-of-fit testing ($n^{-4s/(4s + d)}$). We also show that PCR-LE is \emph{manifold adaptive}: that is, we consider the situation where the design is supported on a manifold of small intrinsic dimension $m$, and give upper bounds establishing that PCR-LE achieves the faster minimax estimation ($n^{-2s/(2s + m)}$) and testing ($n^{-4s/(4s + m)}$) rates of convergence. Interestingly, these rates are almost always much faster than the known rates of convergence of graph Laplacian eigenvectors to their population-level limits; in other words, for this problem regression with estimated features appears to be much easier, statistically speaking, than estimating the features itself. We support these theoretical results with empirical evidence.
\end{abstract}

### 2021-12-22 manuscripts/laplacian_eigenmaps/main_text.tex
Added cover letter for Inf and Inf.

Laplacian Eigenmaps (LE)~\citep{belkin03a} is a method for nonlinear dimensionality reduction and data representation. Given data points $\{X_1,\ldots,X_n\} \subset \Reals^d$, LE maps each $X_i$ to a vector $(v_{1,i},\ldots,v_{K,i})$ according to the following steps.
\begin{enumerate}
	\item First, LE forms a \emph{neighborhood graph} $G = (V,W)$ over the points $\{X_1,\ldots,X_n\}$. The graph $G$ is an undirected, weighted graph, with vertices $V = \{X_1,\ldots,X_n\}$, and weighted edges $W_{ij}$ which correspond to the proximity between points $X_i$ and $X_j$.
	\item Next, LE forms an (unweighted) \emph{graph Laplacian} matrix $L \in \Reals^{n \times n}$, a symmetric and diagonally dominant matrix with diagonal elements $L_{ii} = \sum_{j = 1}^{n} W_{ij}$, and off-diagonal elements $L_{ij} = -W_{ij}$. 
	\item Finally, LE takes the eigendecomposition $L = \sum_{k = 1}^{n} \lambda_k v_k v_k^{\top}$, and outputs the vectors $(v_{1,i},\ldots,v_{K,i}) \in \Reals^K$ for each $i = 1,\ldots,n$.
\end{enumerate} 
A natural way to use LE is by taking the vectors $\{(v_{1,i},\ldots,v_{K,i})\}_{i = 1}^{n}$ to be features in a downstream regression algorithm. In this paper, we study a simple method along these lines: Principal Components Regression with Laplacian-Eigenmaps (PCR-LE), a method for nonparametric regression which operates by running ordinary least squares (OLS) using the features output by LE. Given pairs of design points and responses $(X_1,Y_1),\ldots, (X_n,Y_n)$, PCR-LE computes an estimate $\wh{f} \in \Reals^n$,
\begin{equation}
\label{eqn:pcr-le}
\wh{f} := \argmin_{f \in \mathrm{span}\{v_1,\ldots,v_K\}} \|{\bf Y} - f\|_2^2
\end{equation}
where ${\bf Y} = (Y_1,\ldots,Y_n) \in \Reals^n$ is the vector of responses and $\|\cdot\|_2$ denotes the usual Euclidean norm in $\Reals^n$. (For a formal definition of LE and PCR-LE, see Section~\ref{subsec:laplacian_eigenmaps}.)

@@

We also view PCR-LE as being particularly well-suited for regression over Sobolev spaces due to their close connection with \emph{spectral series regression}. Spectral series regression computes generalized empirical Fourier coefficients $\wt{a}_k := \frac{1}{n}\sum_{i = 1}^{n} Y_i \psi_k(X_i)$, and truncates to the $K$-lowest frequency eigenfunctions of $\Delta_P$, producing the estimate
\begin{equation}
\label{eqn:population-level_spectral_series}
\wt{f}(x) = \sum_{k = 1}^{K} \wt{a}_k \psi_k(x).
\end{equation} 
Spectral series regression is intrinsically linked with Sobolev spaces. That is because under appropriate boundary conditions, a unit ball in the order-$s$ Sobolev space consists of functions $f = \sum_{k} a_k \psi_k \in L^2(\mc{X})$ for which the generalized Fourier coefficients $\{a_k\}_{k = 1}^{\infty}$ satisfy the decay condition $\sum_{k} a_k^2 \rho_k^s \leq C$ (See Section~\ref{subsec:spectral_projection} for more details). This decay condition justifies the truncated series estimator \eqref{eqn:population-level_spectral_series}, since it means the truncation will incur only a limited amount of bias for any $f_0 \in H^s(\mc{X})$. For this reason spectral series regression over Sobolev spaces has been well-studied---at least when $\mc{X} = [0,1]^d$---since at least \citet{rice1984},\footnote{And proposed much earlier in the context of density estimation by \citet{cencov1962}.} and its optimality properties are by this point generally well-understood.

PCR-LE serves as an empirical approximation to spectral series regression, since as already mentioned the eigenvectors $v_k$ are empirical approximations to the eigenfunctions $\psi_k$ of $\Delta_P$. Viewed in this light, a major advantage of PCR-LE is that it operates without needing knowledge of the design distribution $P$. This is an advantage because in our context $P$ is treated as an unknown and potentially complex distribution, and for example can be highly non-uniform, have a complicated support which may be a submanifold of $\Rd$, or both. In contrast, spectral series regression relies on diagonalizing the density-weighted Laplacian $\Delta_P$, and in our context must be viewed as an oracle method; to emphasize this we henceforth refer to the estimator defined in~\eqref{eqn:population-level_spectral_series} as \emph{population-level spectral series regression}. On the other hand, intuitively PCR-LE incurs some extra error by using an empirical approximation to the underlying basis $\{\psi_k\}_{k = 1}^{\infty}$: our work shows that in many cases, this extra error is not enough to change the overall rate of convergence.

@@

\paragraph{Perspective: Regression with Empirical Features vs. Feature Estimation.}
We now pause for a moment, to emphasize that in a certain respect the aforementioned rates of convergence for PCR-LE are quite surprising. Remember that PCR-LE is a regression method using features (eigenvectors $v_k$ of the graph Laplacian $L$) which are themselves empirical estimates of population-level quantities (eigenfunctions $\psi_k$ of the density-weighted Laplacian $\Delta_P$). It seems reasonable to expect that the error of PCR-LE should be decomposed into two parts: first, the error with which these empirically-derived features estimate their continuum limits; second, the error with which, given ideal population-level features, the regression function is learned.

Crucially, our analysis \emph{does not} work in this way. This is important because all known upper bounds on the rates at which $v_k \to \psi_K$ as $n \to \infty$ are much slower than the minimax rates for regression over Sobolev classes. For instance, the best currently known upper bound on the empirical $L^2$ error $\frac{1}{n}\sum_{i = 1}^{n}(\sqrt{n} v_{k,i} - \psi_k(X_i))^2$ is only on the order of $n^{-2/(4 + d)}$~\cite{cheng2021}, which is slower than the minimax estimation rate over $H^s(\mc{X})$ for any $s \in \mathbb{N}, s \geq 1$.\footnote{To make matters worse, PCR-LE, when deployed optimally, does not use a single eigenvector $v_k$ for a fixed index $k \in \mathbb{N}$, but rather many eigenvectors $v_1,\ldots,v_K$ with $K$ growing in $n$. As $K$ grows larger, the rate at which $v_K \to \psi_K$ gets slower, since the population-level object being estimated is less regular; see~\citep{burago2014,trillos2019}.} Although this upper bound may not reflect the true rate of convergence of graph Laplacian eigenvectors---this is still an active area of research, and no lower bounds are known---it seems very unlikely that the true rate matches the minimax estimation rate $n^{-2s/(2s + d)}$, which after all approaches the dimension-free rate $1/n$ for large values of $s$. The bottom line is that the rate at which graph Laplacian eigenvectors are known to converge to density-weighted Laplacian eigenfunctions is too slow to explain the upper bounds we establish for PCR-LE.

@@

Ultimately our upper bound on the error of PCR-LE is determined entirely by a pair of graph functionals: the quadratic form $f_0^{\top}L^s f_0$, and the graph Laplacian eigenvalue $\lambda_{K + 1}$. This brings a couple of advantages. First, it eliminates the need to analyze convergence of eigenvectors to eigenfunctions, which is critical in order to get sufficiently fast rates of convergence for PCR-LE, as we have already explained. Instead, we only have to consider these two graph functionals, both of which are known to converge at faster rates than graph Laplacian eigenvectors. Second, in order to obtain upper bounds on $\|\wh{f} - f_0\|_n^2$ we do not require that these graph functionals themselves converge to population-level limits, but only that they be stochastically bounded on the right order, which is a much weaker requirement. To derive our upper bounds on the error of PCR-LE, we directly analyze the quadratic form $f_0^{\top}L^s f_0$ and the eigenvalue $\lambda_{K + 1}$, using some existing results as well as deriving some new ones which may be of independent interest.

@@

In this supervised setting, there has been relatively little work regarding \emph{random design} regression with neighborhood graph Laplacians . Aside from our own work on Laplacian smoothing, summarized above, we highlight two other related papers: \citet{lee2016}, who analyze a variant of PCR-LE, but derive suboptimal rates of convergence, and \citet{trillos2020}, who study Laplacian smoothing and establish the uniform upper bound $\max_{i = 1,\ldots,n}|\wc{f}(X_i) - f_0(X_i)| \leq C n^{-2/(2 + d)}$ under the assumption $f_0 \in C^2(\mc{X})$, which is slower than the minimax rate $n^{-4/(4 + d)}$ for this class.

@@

\subsection{Spectral series regression over Sobolev spaces}
\label{subsec:spectral_projection}
We now establish some upper bounds on the error of population-level spectral series regression when $f_0 \in H^s(\mc{X})$, which imply that such methods achieve optimal rates of convergence for both estimation and testing. The upper bounds we establish are ``usual'' in the sense that they match the rates $n^{-2s/(2s + d)}$ (estimation) and $n^{-4s/(4s + d)}$ (testing) which are already known in many cases. However, they are unusual in that we treat both the case where $s < d/2$ and thus the Sobolev space $H^s(\mc{X})$ does not continuously embed into $C(\mc{X})$, and the case where $P$ is not the uniform distribution over the unit cube. The upper bounds given in this section serve two purposes: first, to clarify what the rates are in these less-typically studied settings; second, to show that population-level spectral series regression can obtain the optimal rate. This latter point is important since the method we focus on for the most part, PCR-LE, is an empirical approximation to population-level spectral series regression.

@@

For the signal detection problem, the population-level spectral series test $\wt{\varphi} := \1\{\wt{T} \geq K/N + \sqrt{2K/an^2}\}$ has bounded Type I error, $\Ebb_{0}[\wt{\varphi}] \leq a (1 + o(1))$. Proposition~\ref{prop:spectral_series_testing} gives an upper bound on the Type II error that holds over all $f_0 \in \mc{H}^s(\mc{X};M)$ for which $\|f_0\|_P^2$ is sufficiently large.
\begin{proposition}
	\label{prop:spectral_series_testing}
	Suppose data is observed according to Model~\ref{def:model_flat_euclidean}, and that the density $p$ is known.  Suppose additionally that $\partial \mc{X} \in C^{1,1}$, $p \in C^{\infty}(\mc{X})$, $f_0 \in \mc{H}^{s}(\mc{X};M)$ for some $s > d/4$, and $\|f_0\|_{L^4(\mc{X})}^4 \leq 1$. Then there exists a constant $C$ which does not depend on $f_0$, $M$ or $n$ such that the following statement holds: if the population-level spectral series test $\wt{\varphi}$ is computed with parameter $K = \max\{\floor{M^2n}^{2d/(4s + d)},1\}$, and if
	\begin{equation}
	\label{eqn:spectral_series_testing}
	\|f_0\|_P^2 \geq C\min\Bigl\{M^2(M^2n)^{-4s/(4s + d)}, \frac{1}{n}\Bigr\}
	\end{equation}
	then the Type II error is upper bounded, $\Ebb_{f_0}[1 - \phi] \leq b$.
\end{proposition}
Assuming again that $n^{-1/2} \lesssim M$, the upper bound in~\eqref{eqn:spectral_series_testing} is $M^2(M^2n)^{-4s/(4s + d)}$, matching the usual minimax critical radius over Sobolev spaces (see e.g.~\cite{guerre02,ingster2009,ingster2012}. Specifically, \cite{ingster2009} show that the minimax squared critical radius is on the order of $n^{-4s/(4s + d)}$ when $M = 1$, and simple alterations of their analysis imply the rate $M^2(M^2n)^{-4s/(4s + d)}$ for general $M$.)

On the other hand, when $s \leq d/4$ the minimax regression testing rates over $H^s(\mc{X})$ are not known. If one explicitly assumes $f_0 \in L^4(\mc{X};1)$---note that $H^s(\mc{X})$ does not continuously embed into $L^4(\mc{X})$ when $s \leq d/4$---then the minimax critical radius for regression testing is on the order of $n^{-1/2}$ \citep{guerre02}, and is achieved by a test using the naive statistic $\|{\bf Y}\|_n^2$. In other words, the regression testing problem over Sobolev spaces fundamentally changes when $s \leq d/4$, and hereafter when we discuss testing we will limit our consideration to $s > d/4$.

@@

	\item \emph{In-sample error.} Since the PCR-LE estimator is defined only at the design points $\{X_1,\ldots,X_n\}$, we use the empirical norm $\|\cdot\|_n^2$ as our estimation loss. Depending on the problem at hand, it may be more interesting to consider loss in $L^2(P)$ norm, for instance because this loss is intrinsically tied to the prediction error $\Ebb_{X \sim P}(\wh{f}(X) - f_0(X))^2$. In some preliminary analysis, we have considering applying a generic kernel smoother $T$ to the PCR-LE estimate, so as to produce a bona-fide function $T\wh{f}: \Rd \to \Reals$; the error $\|T\wh{f} - f_0\|_P^2$ is provably on the order of the minimax estimation rate $n^{-2/(2 + d)}$. Indeed, our analysis is completely independent of the structure of $\wh{f}$, and could apply to any estimator $\wh{g} \in \Reals^n$ sufficiently close to $f_0$ in empirical norm. We intend to pursue the matter further in future work.

@@

\begin{enumerate}[label=(P\arabic*)]
	\setcounter{enumi}{1}
	\item 
	\label{asmp:parameters_testing_fo}
	The graph radius $\varepsilon$ satisfies~\eqref{eqn:radius_fo}, and the number of eigenvectors 
	\begin{equation}
	\label{eqn:eigenvector_testing_fo}
	K = \min\Bigl\{\floor{(M^2n)^{2d/(4 + d)}} \vee 1, n\Bigr\}.
	\end{equation}
\end{enumerate}
\begin{theorem}
	\label{thm:laplacian_eigenmaps_testing_fo}
	Fix $a,b \in (0,1)$. Suppose Model~\ref{def:model_flat_euclidean}. Then $\mathbb{E}_0[\varphi] \leq a$, i.e $\varphi$ is a level-$a$ test. Suppose additionally $f_0 \in H^1(\mc{X};M)$, and that $d < 4$. Then there exist constants $C$ and $N$ that do not depend on $f_0$, such that the following statement holds for all $n \geq N$: if the PCR-LE test $\varphi$ is computed with a kernel $\eta$ satisfying~\ref{asmp:kernel_flat_euclidean}, and parameters $\varepsilon$ and $K$ satisfying~\ref{asmp:parameters_testing_fo}, and if $f_0$ satisfies
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_testing_criticalradius_fo}
	\|f_0\|_P^2 \geq C\biggl(\Bigl(M^2(M^2n)^{-4/(4 + d) } \wedge n^{-1/2}\Bigr)\biggl[\sqrt{\frac{1}{a}} + \frac{1}{b}\biggr] \vee \frac{M^2}{b n^{2/d}} \biggr) \vee \frac{1}{n},
	\end{equation}
	then $\Ebb_{f_0}[1 - \phi] \leq b$.
\end{theorem}
Although~\eqref{eqn:laplacian_eigenmaps_testing_criticalradius_fo} involves taking the maximum of several different terms, the important takeaway of Theorem~\ref{thm:laplacian_eigenmaps_testing_fo} is that if $n^{-1/2} \lesssim M \lesssim n^{(4 - d)/4d}$, then $\varphi$ has small worst-case risk as long as $f_0$ is separated from $0$ by at least $M^2(M^2n)^{-4/(4 + d)}$. This implies that $\varphi$ is a minimax rate-optimal test over $H^1(\mc{X};M)$ when $d \in \{1,2,3\}$. As mentioned previously, when $d \geq 4$ the first order Sobolev space $H^1(\mc{X})$ does not continuously embed into $L^4(\mc{X})$, and in this case the optimal rates for regression testing over Sobolev spaces are unknown.

@@

At a high level, it is intuitively reasonable that PCR-LE should have more difficulty achieving the minimax rates of convergence for testing, as opposed to estimation. To obtain the faster rates of convergence for testing, PCR-LE must use many more eigenvectors than are necessary for estimation, including some which correspond to very large eigenvalues. It is known that the approximation properties of eigenvectors corresponding to large eigenvalues are very poor~\citep{burago2014,trillos2019}, and when $d > 4$ this prevents us from establishing that PCR-LE is an optimal test. That being said, although we suspect $\varphi$ is truly suboptimal when $d > 4$, our analysis relies on an upper bound on testing bias. Since we do not prove a matching lower bound, we cannot rule out that the test $\varphi$ is optimal for all $s < d/4$. We leave the matter to future work.

>>>>

Laplacian Eigenmaps (LE)~\citep{belkin03a} is a method for nonlinear dimensionality reduction and data representation. Given data points $\{X_1,\ldots,X_n\} \subset \Reals^d$, LE maps each $X_i$ to a vector $(v_{1,i},\ldots,v_{K,i})$ according to the following steps.
\begin{enumerate}
	\item First, LE forms a \emph{neighborhood graph} $G = (V,W)$ over the points $\{X_1,\ldots,X_n\}$. The graph $G$ is an undirected, weighted graph, with vertices $V = \{X_1,\ldots,X_n\}$, and weighted edges $W_{ij}$ which correspond to the proximity between points $X_i$ and $X_j$.
	\item Next, LE forms an (unweighted) \emph{graph Laplacian} matrix $L \in \Reals^{n \times n}$, a symmetric and diagonally dominant matrix with diagonal elements $L_{ii} = \sum_{j = 1}^{n} W_{ij}$, and off-diagonal elements $L_{ij} = -W_{ij}$. 
	\item Finally, LE takes the eigendecomposition $L = \sum_{k = 1}^{n} \lambda_k v_k v_k^{\top}$, and outputs the vectors $(v_{1,i},\ldots,v_{K,i}) \in \Reals^K$ for each $i = 1,\ldots,n$.
\end{enumerate} 
A natural way to use LE is by taking the collection of vectors $\{(v_{1,i},\ldots,v_{K,i})\}_{i = 1}^{n}$ to be features in a downstream regression algorithm. In this paper, we study a simple method along these lines: Principal Components Regression with Laplacian-Eigenmaps (PCR-LE), a method for nonparametric regression which operates by running ordinary least squares (OLS) using the features output by LE. Given pairs of design points and responses $(X_1,Y_1),\ldots, (X_n,Y_n)$, PCR-LE computes an estimate $\wh{f} \in \Reals^n$,
\begin{equation}
\label{eqn:pcr-le}
\wh{f} := \argmin_{f \in \mathrm{span}\{v_1,\ldots,v_K\}} \|{\bf Y} - f\|_2^2
\end{equation}
where ${\bf Y} = (Y_1,\ldots,Y_n) \in \Reals^n$ is the vector of responses and $\|\cdot\|_2$ denotes the usual Euclidean norm in $\Reals^n$. (For a formal definition of LE and PCR-LE, see Section~\ref{subsec:laplacian_eigenmaps}.)

@@

We also view PCR-LE as being particularly well-suited for regression over Sobolev spaces due to their close connection with \emph{spectral series regression}. Spectral series regression computes generalized empirical Fourier coefficients $\wt{a}_k := \frac{1}{n}\sum_{i = 1}^{n} Y_i \psi_k(X_i)$, and truncates to the $K$-lowest frequency eigenfunctions of $\Delta_P$, producing the estimate
\begin{equation}
\label{eqn:population-level_spectral_series}
\wt{f}(x) = \sum_{k = 1}^{K} \wt{a}_k \psi_k(x).
\end{equation} 
Spectral series regression is intrinsically linked with Sobolev spaces. That is because under appropriate boundary conditions, a ball in the order-$s$ Sobolev space consists of functions $f = \sum_{k} a_k \psi_k \in L^2(\mc{X})$ for which the generalized Fourier coefficients $\{a_k\}_{k = 1}^{\infty}$ satisfy the decay condition $\sum_{k} a_k^2 \rho_k^s \leq C$ (See Section~\ref{subsec:spectral_projection} for more details). This decay condition justifies the truncated series estimator \eqref{eqn:population-level_spectral_series}, since it means the truncation will incur only a limited amount of bias for any $f_0 \in H^s(\mc{X})$. For this reason spectral series regression over Sobolev spaces has been well-studied---at least when $\mc{X} = [0,1]^d$---since at least \citet{rice1984},\footnote{And proposed much earlier in the context of density estimation by \citet{cencov1962}.} and its optimality properties are by this point generally well-understood.

PCR-LE serves as an empirical approximation to spectral series regression, since as already mentioned the eigenvectors $v_k$ are empirical approximations to the eigenfunctions $\psi_k$ of $\Delta_P$. Viewed in this light, a major advantage of PCR-LE is that it operates without needing knowledge of the design distribution $P$. This is an advantage because in our context $P$ is an unknown and potentially complex distribution: for example, it can be highly non-uniform, have a complicated support which may be a submanifold of $\Rd$, or both. In contrast, spectral series regression relies on diagonalizing the density-weighted Laplacian $\Delta_P$, and in our context must be viewed as an oracle method; to emphasize this we henceforth refer to the estimator defined in~\eqref{eqn:population-level_spectral_series} as \emph{population-level spectral series regression}. On the other hand, intuitively PCR-LE incurs some extra error by using an empirical approximation to the underlying basis $\{\psi_k\}_{k = 1}^{\infty}$: our work shows that in many cases, this extra error is not enough to change the overall rate of convergence.

@@

\paragraph{Perspective: regression error versus feature reconstruction.}
We now pause for a moment, to emphasize that in a certain respect the aforementioned rates of convergence for PCR-LE are quite surprising. Remember that PCR-LE is a regression method using features (eigenvectors $v_k$ of the graph Laplacian $L$) which are themselves empirical estimates of population-level quantities (eigenfunctions $\psi_k$ of the density-weighted Laplacian $\Delta_P$). It seems reasonable to expect that the error of PCR-LE should be decomposed into two parts: first, the error with which these empirically-derived features estimate their continuum limits; second, the error with which, given ideal population-level features, the regression function is learned.

Crucially, our analysis \emph{does not} work in this way. This is important because all known upper bounds on the rates at which $v_k \to \psi_k$ as $n \to \infty$ are much slower than the minimax rates for regression over Sobolev classes. For instance, the best currently known upper bound on the empirical $L^2$ error $\frac{1}{n}\sum_{i = 1}^{n}(\sqrt{n} v_{k,i} - \psi_k(X_i))^2$ is only on the order of $n^{-2/(4 + d)}$~\citep{cheng2021}, which is slower than the minimax estimation rate over $H^s(\mc{X})$ for any $s \in \mathbb{N}, s \geq 1$.\footnote{To make matters worse, PCR-LE, when deployed optimally, does not use a single eigenvector $v_k$ for a fixed index $k \in \mathbb{N}$, but rather many eigenvectors $v_1,\ldots,v_K$ with $K$ growing in $n$. As $K$ grows larger, the rate at which $v_K \to \psi_K$ gets slower, since the population-level object being estimated is less regular; see~\citep{burago2014,trillos2019}.} Although this upper bound may not reflect the true rate of convergence of graph Laplacian eigenvectors---this is still an active area of research, and no lower bounds are known---it seems very unlikely that the true rate matches the minimax estimation rate $n^{-2s/(2s + d)}$, which after all approaches the dimension-free rate $1/n$ for large values of $s$. The bottom line is that the rate at which graph Laplacian eigenvectors are known to converge to density-weighted Laplacian eigenfunctions is too slow to explain the upper bounds we establish for PCR-LE.

@@

Ultimately our upper bound on the error of PCR-LE is determined entirely by a pair of graph functionals: the quadratic form $f_0^{\top}L^s f_0$, and the graph Laplacian eigenvalue $\lambda_{K + 1}$. This brings a couple of advantages:
\begin{itemize}
	\item First, it eliminates the need to analyze convergence of eigenvectors to eigenfunctions, which is critical in order to get sufficiently fast rates of convergence for PCR-LE, as we have already explained. Instead, we only have to consider these two graph functionals, both of which are known to converge at faster rates than graph Laplacian eigenvectors. 
	\item Second, in order to obtain upper bounds on $\|\wh{f} - f_0\|_n^2$ we do not require that these graph functionals themselves converge to population-level limits, but only that they be stochastically bounded on the right order. The latter is a much weaker requirement. 
\end{itemize}
To derive our upper bounds on the error of PCR-LE, we directly analyze the quadratic form $f_0^{\top}L^s f_0$ and the eigenvalue $\lambda_{K + 1}$, using some existing results as well as deriving some new ones which may be of independent interest.

@@

In this supervised setting, there has been relatively little work regarding \emph{random design} regression with neighborhood graph Laplacians . Aside from our own work on Laplacian smoothing, summarized above, we highlight two other related papers: \citet{lee2016}, who analyze a variant of PCR-LE, but derive suboptimal rates of convergence, and \citet{trillos2020}, who study Laplacian smoothing and establish the uniform upper bound $\max_{i = 1,\ldots,n}|\wc{f}(X_i) - f_0(X_i)| \leq C n^{-2/(2 + d)}$ under the assumption $f_0 \in C^2(\mc{X})$, which is slower than the minimax rate $n^{-4/(4 + d)}$ for this function class.

@@

\subsection{Spectral series regression over Sobolev spaces}
\label{subsec:spectral_projection}
We now establish some upper bounds on the error of population-level spectral series regression when $f_0 \in H^s(\mc{X})$, which imply that such methods achieve optimal rates of convergence for both estimation and testing. The upper bounds we establish are ``usual'' in the sense that they match the rates $n^{-2s/(2s + d)}$ (estimation) and $n^{-4s/(4s + d)}$ (testing) which are already known in many cases. However, they are unusual in that we treat both the case where $s < d/2$ and thus the Sobolev space $H^s(\mc{X})$ does not continuously embed into $C(\mc{X})$, and the case where $P$ is not the uniform distribution over the unit cube. The upper bounds given in this section serve two purposes: first, to clarify what the rates are in these less-typically studied settings; second, to show that even in this general setting, population-level spectral series regression can always obtain the optimal rates. This latter point is important since the method we focus on for the most part, PCR-LE, is an empirical approximation to population-level spectral series regression.

@@

For the signal detection problem, the population-level spectral series test $\wt{\varphi} := \1\{\wt{T} \geq K/N + \sqrt{2K/an^2}\}$ has bounded Type I error, $\Ebb_{0}[\wt{\varphi}] \leq a (1 + o(1))$. Proposition~\ref{prop:spectral_series_testing} gives an upper bound on the Type II error that holds over all $f_0 \in \mc{H}^s(\mc{X};M)$ for which $\|f_0\|_P^2$ is sufficiently large.
\begin{proposition}
	\label{prop:spectral_series_testing}
	Suppose data is observed according to Model~\ref{def:model_flat_euclidean}, and that the density $p$ is known.  Suppose additionally that $\partial \mc{X} \in C^{1,1}$, $p \in C^{\infty}(\mc{X})$, $f_0 \in \mc{H}^{s}(\mc{X};M)$ for some $s > d/4$, and $\|f_0\|_{L^4(\mc{X})}^4 \leq 1$. Then there exists a constant $C$ which does not depend on $f_0$, $M$ or $n$ such that the following statement holds: if the population-level spectral series test $\wt{\varphi}$ is computed with parameter $K = \max\{\floor{M^2n}^{2d/(4s + d)},1\}$, and if
	\begin{equation}
	\label{eqn:spectral_series_testing}
	\|f_0\|_P^2 \geq C\min\Bigl\{M^2(M^2n)^{-4s/(4s + d)}, \frac{1}{n}\Bigr\}
	\end{equation}
	then the Type II error is upper bounded, $\Ebb_{f_0}[1 - \wt{\varphi}] \leq b$.
\end{proposition}
Assuming again that $n^{-1/2} \lesssim M$, the right hand side of~\eqref{eqn:spectral_series_testing} is $M^2(M^2n)^{-4s/(4s + d)}$, matching the usual minimax critical radius over Sobolev spaces (see e.g.~\cite{guerre02,ingster2009,ingster2012}. Specifically, \cite{ingster2009} show that the minimax squared critical radius is on the order of $n^{-4s/(4s + d)}$ when $M = 1$, and simple alterations of their analysis imply the rate $M^2(M^2n)^{-4s/(4s + d)}$ for general $M$.) On the other hand, when $s \leq d/4$ the minimax regression testing rates over $H^s(\mc{X})$ are not known. If one explicitly assumes $f_0 \in L^4(\mc{X};1)$---note that $H^s(\mc{X})$ does not continuously embed into $L^4(\mc{X})$ when $s \leq d/4$---then the minimax critical radius for regression testing is on the order of $n^{-1/2}$ \citep{guerre02}, and is achieved by a test using the naive statistic $\|{\bf Y}\|_n^2$. In other words, the regression testing problem over Sobolev spaces fundamentally changes when $s \leq d/4$, and hereafter when we discuss testing we will limit our consideration to $s > d/4$.

@@

	\item \emph{In-sample error.} Since the PCR-LE estimator is defined only at the design points $\{X_1,\ldots,X_n\}$, we use the empirical norm $\|\cdot\|_n^2$ as our estimation loss. Depending on the problem at hand, it may be more interesting to consider loss in $L^2(P)$ norm, for instance because this loss is intrinsically tied to the prediction error $\Ebb_{X \sim P}(\wh{f}(X) - f_0(X))^2$. In some preliminary analysis, we have considered applying a generic kernel smoother $T$ to the PCR-LE estimate, so as to produce a bona-fide function $T\wh{f}: \Rd \to \Reals$; the error $\|T\wh{f} - f_0\|_P^2$ is provably on the order of the minimax estimation rate $n^{-2/(2 + d)}$. Indeed, our analysis is completely independent of the structure of $\wh{f}$, and could apply to any estimator $\wh{g} \in \Reals^n$ sufficiently close to $f_0$ in empirical norm. We intend to pursue the matter further in future work.

@@

\begin{enumerate}[label=(P\arabic*)]
	\setcounter{enumi}{1}
	\item 
	\label{asmp:parameters_testing_fo}
	The graph radius $\varepsilon$ satisfies~\eqref{eqn:radius_fo}, and the number of eigenvectors 
	\begin{equation}
	\label{eqn:eigenvector_testing_fo}
	K = \min\Bigl\{\floor{(M^2n)^{2d/(4 + d)}} \vee 1, n\Bigr\}.
	\end{equation}
\end{enumerate}
\begin{theorem}
	\label{thm:laplacian_eigenmaps_testing_fo}
	Fix $a,b \in (0,1)$. Suppose Model~\ref{def:model_flat_euclidean}. Then $\mathbb{E}_0[\varphi] \leq a$, i.e $\varphi$ is a level-$a$ test. Suppose additionally $f_0 \in H^1(\mc{X};M)$, and that $d < 4$. Then there exist constants $C$ and $N$ that do not depend on $f_0$, such that the following statement holds for all $n \geq N$: if the PCR-LE test $\varphi$ is computed with a kernel $\eta$ satisfying~\ref{asmp:kernel_flat_euclidean}, and parameters $\varepsilon$ and $K$ satisfying~\ref{asmp:parameters_testing_fo}, and if $f_0$ satisfies
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_testing_criticalradius_fo}
	\|f_0\|_P^2 \geq C\biggl(\Bigl(M^2(M^2n)^{-4/(4 + d) } \wedge n^{-1/2}\Bigr)\biggl[\sqrt{\frac{1}{a}} + \frac{1}{b}\biggr] \vee \frac{M^2}{b n^{2/d}} \biggr) \vee \frac{1}{n},
	\end{equation}
	then $\Ebb_{f_0}[1 - \varphi] \leq b$.
\end{theorem}
Although~\eqref{eqn:laplacian_eigenmaps_testing_criticalradius_fo} involves taking the maximum of several different terms, the important takeaway of Theorem~\ref{thm:laplacian_eigenmaps_testing_fo} is that if $n^{-1/2} \lesssim M \lesssim n^{(4 - d)/4d}$, then $\varphi$ has small worst-case risk as long as $f_0$ is separated from $0$ by at least $M^2(M^2n)^{-4/(4 + d)}$. This implies that $\varphi$ is a minimax rate-optimal test over $H^1(\mc{X};M)$ when $d \in \{1,2,3\}$. As mentioned previously, when $d \geq 4$ the first order Sobolev space $H^1(\mc{X})$ does not continuously embed into $L^4(\mc{X})$, and in this case the optimal rates for regression testing over Sobolev spaces are unknown.

@@

At a high level, it is intuitively reasonable that PCR-LE should have more difficulty achieving the minimax rates of convergence for testing, as opposed to estimation. To obtain the faster rates of convergence for testing, PCR-LE must use many more eigenvectors than are necessary for estimation, including some eigenvectors which correspond to very large eigenvalues. It is known that the approximation properties of eigenvectors corresponding to large eigenvalues are very poor~\citep{burago2014,trillos2019}, and when $d > 4$ this prevents us from establishing that PCR-LE is an optimal test. That being said, although we suspect $\varphi$ is truly suboptimal when $d > 4$, our analysis relies on an upper bound on testing bias. Since we do not prove a matching lower bound, we cannot rule out that the test $\varphi$ is optimal for all $s < d/4$. We leave the matter to future work.
